---
title: "Data Vizualization III Exploratory Analysis"
output: github_document
---

# Setup

```{r setuo}
library(patchwork)
library(janitor)
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() +
  theme(legend.position = "bottom"))


options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Load the weather data

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>% 
  select(name, id, everything())
```

## 'group_by'

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  ungroup(month)
```

## counting things

count month observations

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarise(n_obs = n())

weather_df %>% 
  group_by(name) %>% 
  summarise(n_obs = n())

weather_df %>% 
  group_by(name, month) %>% 
  summarise(n_obs = n())
```

we can 'count'

```{r}
weather_df %>% 
  count(name, month, name = "n_obs")
```

** NEVER use base R's 'table' 

```{r}
weather_df %>% 
  pull(month) %>% 
  table()
```

other helpful counters 

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarise(
    n_obs = n(),
    n_days = n_distinct(date))
```



## digression on 2x2 tables

```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE ~ ""
    )
  ) %>% 
  group_by(name, cold) %>% 
  summarize(count = n())
  
```




```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE ~ ""
    )
  ) %>% 
  janitor::tabyl(name, cold)
```



## general summaries

you can do lots of sum maries

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )
```

this is a dataframe! so we can make a plot

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```


Suppose you want to summarise many columns

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(across(prcp:tmin, mean))
```


Reminder: sometimes your results are easeir to read in another format

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>% 
  knitr::kable(digits = 1)
```

## 'group_by' and 'mutate'



```{r}
weather_df %>% 
  group_by(name) %>%
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```


what about window functions

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  mutate(temp_rank = min_rank(tmax)) %>% 
  filter(temp_rank == 1)
```


lag

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(lag_temp = lag(tmax))

```

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(temp_change = tmax - lag(tmax)) %>% 
  summarize(
    temp_change_max = max(temp_change, na.rm = TRUE),
    temp_change_sd = sd(temp_change, na.rm = TRUE)
  )

```

## quick note

summarize only gets you so far


```{r}
nyc_airbnb = read_csv("./data/nyc_airbnb.csv")

nyc_airbnb = 
  nyc_airbnb %>% 
  mutate(stars = review_scores_location / 2) %>% 
  rename(
    borough = neighbourhood_group,
    neighborhood = neighbourhood) %>% 
  select(price, stars, borough, neighborhood, room_type)
```

```{r}
nyc_airbnb %>% 
  tabyl(borough, room_type)
```

